---
format: html
---

## Repo Overview

This repo explores two classic statistical models for survival analysis using patients who suffered from heart attacks.

Models:

-   Kaplan-Meier
-   Cox Proportional Hazards

```{r}
#| include: false
library(tidyverse)
library(survival)
library(StepReg)
library(ggplot2)
library(scales)
library(dplyr)
library(ggfortify)
```

## Censoring

Right censoring occurs when the event of interest is yet to be observed for some data. Examples include time to death for patients with a terminal illness, time to relapse for patients in rehab, or time to service termination for customers. Here are three hypothetical data points.

```{r}
#| echo: false
tibble(Time = c(1, 2, 3), Event = c(0, 1, 1))
```

The average of time is not two. Due to the first observation being censored, the average is **at least** two. Unless all data points are uncensored, this "at least" affects all summary statistics, hypothesis tests, and models. Survival analysis is the area of statistics designed to deal with this notion of "at least" inherent in the data.

## Data Overview

The data set includes 500 patients who suffered a heart attack.

Variables:

-   id - Identification Code (1 - 500)
-   age - Age at Hospital Admission (years)
-   gender - Gender (0 = Male, 1 = Female)
-   hr - Initial Heart Rate (Beats per minute)
-   sysbp - Initial Systolic Blood Pressure (mmHg)
-   diasbp - Initial Diastolic Blood (mmHg)
-   bmi - Body Mass Index (kg/m\^2)
-   cvd - History of Cardiovascular Disease (0 = No, 1 = Yes)
-   afb - Atrial Fibrillation (0 = No, 1 = Yes)
-   sho - Cardiogenic Shock (0 = No, 1 = Yes)
-   chf - Congestive Heart Complications (0 = No, 1 = Yes)
-   av3 - Complete Heart Block (0 = No, 1 = Yes)
-   miord - MI Order (0 = First, 1 = Recurrent)
-   mitype - MI Type (0 = non Q-wave, 1 = Q-wave)
-   year - Cohort Year
-   admitdate - Hospital Admission Date
-   disdate - Hospital Discharge Date
-   fdate - Date of last Follow Up
-   los - Length of Hospital Stay
-   dstat - Discharge Status from Hospital (0 = Alive, 1 = Dead)
-   lenfol - Total Length of Follow-up
-   fstat - Vital Status at Last Follow-up (0 = Alive 1 = Dead)

The first ten rows look like

```{r}
#| echo: false
whas500 <- readRDS("data/clean_data.rds")
print(whas500)
```

## Kaplan-Meier Model

The Kaplan-Meier model is a non-parametric model. This model makes no distributional assumptions about survival time. The main output is a survival curve.

```{r}
#| echo: false
km_fit <- survfit(Surv(lenfol, fstat) ~ 1, data=whas500)

temp <- tibble(Time = km_fit$time,Surv = km_fit$surv)

ggplot(temp, aes(x = Time, y = Surv)) + 
  geom_step() + 
  scale_x_continuous(breaks = seq(0, 2400, 400), limit = c(0, 2400)) + 
  scale_y_continuous(breaks = seq(0, 1, .10), labels = percent) + 
  labs(title = "Kaplan-Meier Survival Curve", x = "Time", y = "Proportion of People Alive")

```

According to the curve, 70% of patients lived at least 400 days after being admitted to the hospital.

The KM approach can be used to compare different groups. Is atrial fibrillation associated with worse survival? Visually, the green survival curve is much lower than the red survival curve indicating patients with atrial fibrillation died sooner than patients without it.

```{r}
#| echo: false
whas500_02 <- whas500 %>%
  mutate(afb  = factor(if_else(afb  == 1, "Yes", "No")))
km_fit_02 <- survfit(Surv(lenfol, fstat) ~ afb , data=whas500_02)

temp <- tibble(Time = km_fit_02$time,Surv = km_fit_02$surv, Strata = c(rep("No", 348), rep("Yes", 74)))
ggplot(temp, aes(x = Time, y = Surv, color = Strata)) + 
  geom_step() + 
  scale_x_continuous(breaks = seq(0, 2400, 400), limit = c(0, 2400)) + 
  scale_y_continuous(breaks = seq(0, 1, .10), labels = percent) + 
  labs(title = "Atrial Fibrillation", y = "Proportion of People Alive")
```

Could this difference in survival be attributed to random chance alone? The log rank test shows a p-value below 5% indicating no. Atrial fibrillation is associated with worse survival.

```{r}
#| echo: false
survdiff(Surv(lenfol, fstat) ~ afb, data=whas500_02)
```


The Kaplan-Meier model can be used as a descriptive tool or a inference tool. It cannot be used as a prediction tool.

# Cox Proportional Hazards Model

The Cox Proportional Hazards model can provide predictions for unseen data. This gain comes at the cost of assuming a semi-parametric model. With this model, a new survival curve is calculated. 

```{r}
#| echo: false
# stepwise search
temp <- stepwiseCox(
  formula = Surv(lenfol, fstat) ~ age + gender + hr + sysbp + diasbp + bmi + cvd + afb + sho + chf + av3 + miord + mitype, 
  data = whas500,
  selection = "bidirection",
  select = "AIC"
)
# temp[["Selected Varaibles"]]

cox <- coxph(Surv(lenfol, fstat) ~ age + chf + sho + hr + diasbp + bmi + gender + mitype, data = whas500)
# summary(cox)

cox_fit <- survfit(cox)
temp <- tibble(Time = cox_fit$time,Surv = cox_fit$surv)
ggplot(temp, aes(x = Time, y = Surv)) + 
  geom_step() + 
  scale_x_continuous(breaks = seq(0, 2400, 400), limit = c(0, 2400)) + 
  scale_y_continuous(breaks = seq(0, 1, .10), labels = percent) + 
  labs(title = "Cox Proportional Hazards Survival Curve", y = "Proportion of People Alive")
```


```{r}
#| echo: false
summary(cox)$coefficients
```

According to the model (and ignoring the inflated type I error rates), risk of death increases with:

* Increases in age
* Congestive heart complications
* Cardiogenic shock
* Increases in heart rate

Risk of death decreases with:

* Increases in diastolic blood pressure
* Increases in body mass index
* Being male
* Having a Q-way heart attack instead of a non Q wave.

# Survival Curve Comparison

The two models lead to meaningfully different survival curves. This is due to the difference in model assumptions. The Kaplan-Meier model is nonparametric. The Proportional Hazards model is semi-parametric. 

```{r}
#| echo: false
plot_df <- rbind(
  tibble(Time = km_fit$time,Surv = km_fit$surv,Model = rep("KM",length(km_fit$time))),
  tibble(Time = cox_fit$time,Surv = cox_fit$surv,Model = rep("Cox",length(cox_fit$time))))

ggplot(plot_df, aes(x = Time, y = Surv, color = Model)) + 
  geom_step() + 
  scale_x_continuous(breaks = seq(0, 2400, 400), limit = c(0, 2400)) + 
  scale_y_continuous(breaks = seq(0, 1, .10), labels = percent) + 
  labs(title = "Model Survival Curves", y = "Proportion of People Alive")
```

